---
title: "Rcpp Introduction"
Author: "Stephen Cristiano"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

<!-- cache all Rcpp source chunks in single compilation --> 
```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```


```{Rcpp, eval=FALSE}
#include <RcppArmadillo.h>
// Note #include <Rcpp.h> implied by RcppArmadillo.
using namespace Rcpp;
```

# What is Rcpp

* Rcpp: Seamless integration between <tt> R </tt> and <tt> C++ </tt>
* Extremely simple to connect <tt> C++ </tt> with <tt> R </tt>. <tt> Rcpp </tt>
  Library in <tt> C++ </tt> enables <tt> R </tt>-like syntax and operations on
  imported objects from <tt> R </tt>.
* Maintained by Dirk Eddelbuetter and Romain Francois
* Well supported by Rstudio.
* Supported by <tt> knitr </tt> as well (<tt> rcpp </tt> in code chunk header).
  * See source of this <tt> Rmd </tt> for example how to cache all Rcpp source
    chunks into a single compilation unit.
* Supports any C++ language standard the underlying compiler supports: C++98, C++11, C++14, C++17
  * Packages using Rcpp can deploy standards suppported by R: C++, C++11 and very soon C++14

## Simple examples
Let's begin by examining to very simple examples of calling <tt> C++ </tt>
functions within <tt> R </tt>:

```{Rcpp, echo=TRUE, eval=FALSE}
// [[Rcpp::export]]
int square(int x) {
    return x*x;
}

// [[Rcpp::export]]
int add(int x, int y, int z) {
    int sum = x + y + z;
    return sum;
}
```

Note the <tt> // [[Rcpp::export]] </tt> line allows the function to be 
exported into <tt> R </tt>. Within <tt> R </tt>, I can now call:

```{r}
square(7L)
add(1, 2, 3)
```
It is very straightforward to call the <tt>C++</tt> implemented function
within <tt>R</tt>. The motivation behind <tt>Rcpp</tt> comes when you
run into a computation in <tt>R</tt> that create a performance bottleneck
in your code. By instead writing your function in <tt>C++</tt>, you can
achieve a massive speed boost. <tt>Rcpp</tt> seeks to make the interaction
between <tt>R</tt> and <tt>C++</tt> as seamless and convenient as possible.

## Using cppFunction
It can be even easier to parse C++ code in R
```{r cppfunction}
Rcpp::cppFunction('double accu(NumericVector x) { return(
      std::accumulate(x.begin(), x.end(), 0.0)
   );
}')
res <- accu(seq(1, 10, by=0.5))
res
```
<tt>cppFunction</tt> parses C++ code and compiles it for use within R.
If you have an external .cpp file you want to call within R, can use `sourceCpp`.
You can also evaluate a single C++ statement with <tt>evalCpp()</tt>

```{r evalcpp}
# Showing maximum value of double.
Rcpp::evalCpp('std::numeric_limits<double>::max()')
```


## Why C++?

* One of the most frequently used programming languages in the world.
* Speed.
* Good chance what you want is already implemented in <tt>C++</tt>
* From wikipedia: ‘<tt>C++</tt> is a statically typed, free-form, multi-paradigm,
  compiled, general-purpose, powerful programming language.’

## Why not C++?

* More difficult to debug
* More difficult to modify
* The The population of potentials users who understand both R and C++ is smaller.

## Why Rcpp?
* Easy to use (honest).
* Clean and approachable API that enable for high performance code.
* R style vectorized code at C++ level.
* Programmer time vs computer time: much more efficient code that does not take
  much longer to write.
* Enables access to advanced data structures and algorithms implemented in C++
  but not provided by R.
  * *eg* STL (C++ Standard Template Library), Boost
* Handles garbage collection and the Rcpp programmer should never have to worry
  about memory allocation and deallocation.
* If you are developing a package and you want people to use that package, you
  probably want it to be fast.

# Quick C++ primer
Here's a 2-minute C++ introduction:
```{Rcpp, echo=TRUE, eval=FALSE}
// This is a comment
/* Also this */

// [[Rcpp::export]]
double sumC(NumericVector x) {
    int n = x.size();
    double total = 0;
    for(int i = 0; i < n; ++i) {
        total += x[i]; if(total > 100) break;
    }
    return total;
}
```

```{r}
sumC(seq(1:10))
```

Notice the following things about C++:

* Need to initialize your variables with data type.
* for loops of structure for(initialization; condition; increment).
* Conditionals are the same as R.
* End every statement with a semicolon;
* Vectors and arrays are 0-indexed.
* <tt>size()</tt> is a member function on the vector class - <tt>x.size()</tt>
  returns the size of x. 
* We also saw previously you can call a function from a particular C++ library
  with <tt>::</tt> (<tt>std::accumulate</tt>), similar to R.
* While <tt>C++</tt> can be a very complex language, just knowing these will
  enable you to write faster R functions.

# Rcpp Basics

A quick introduction to Rcpp objects and operations:

## Data structures

* All R objects are internally represented by a SEXP: a pointer to an S expression object.
* Any R object can be passed down to C++ code: vectors, matrices lists. Even functions and environments.
* A large number of user-visible classes for R objects, which contain pointers the the SEXP object.
  * IntegerVector
  * NumericVector
  * LogicalVector
  * CharacterVector
  * NumericMatrix
  * S4
  * and many more

## Rcpp Sugar

* Rcpp sugar brings a higher level of abstraction to C++ code written in Rcpp.
* Avoid C++ loops with code that strongly resembles R.
* Takes advantage of operator overloading.
* Despite the similar syntax, peformance is much faster in C++, though not
  quite as fast as manually optimized C++ code.

Example:
```{Rcpp sugar, eval=FALSE}
// Rcpp implementation
// [[Rcpp::export]]
NumericVector pdistCpp(double x, NumericVector ys) {
    return pow((x-ys), 2);
}
```
```{r sugar2}
# R implementation
pdistR <- function(x, ys) {
    (x - ys)^2
}

pdistR(5.0, c(4.1, -9.3, 0, 13.7))
pdistCpp(5.0, c(4.1, -9.3, 0, 13.7))
```
Note the similarity of the Rcpp and R implementations. Rcpp performs R styled
vectorizations.

## Logical Operators
```{c, eval=FALSE}
// two integer vectors of the same size
NumericVector x;
NumericVector y;
// expressions involving two vectors
LogicalVector res = x < y;
LogicalVector res = x != y;
// one vector, one single value
LogicalVector res = x < 2;
// two expressions
LogicalVector res = (x + y) == (x*x);
// functions producing single boolean result
all(x*x < 3);
any(x*x < 3);
```

## R-like functions
There are many functions similar to what exists inside R:
```{c, eval=FALSE}
is_na(x);
seq_along(x);
sapply( seq_len(10), square<int>() );
ifelse( x < y, x, (x+y)*y );
pmin( x, x*x);
diff( xx );
intersect( xx, yy); //returns interserct of two vectors
unique( xx ); // subset of unique values in input vecto

// math functions
abs(x); exp(x); log(x); ceil(x);
sqrt(x); sin(x); gamma(x);
range(x);
mean(x); sd(x); var(x);
which_min(x); which_max(x);
// A bunch more
```

## Density and RNG functions

Rcpp has access to the same density, distribution, and RNG functions used by R
itself. The seed you set in R will even be passed into Rcpp. For example, you
can draw from a gamma distribution with scale and shape parameters equal to 1 with:

```{Rcpp gamma, eval=FALSE}
RNGScope scope;
// [[Rcpp::export]]
NumericVector getRGamma() {
            NumericVector x = rgamma( 10, 1, 1 );
            return x;
}
```
```{r rgamma}
set.seed(1234)
getRGamma()
set.seed(1234)
rgamma(10, 1, 1)
```


## Printing to console:
Don't use <tt>stdout</tt> and <tt>stderr</tt> as you would in C++. Instead,
use <tt>Rcpp::Rcout</tt> or <tt>Rcpp::print</tt>.

```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
void useOperatorOnVector(NumericVector x) { 
    Rcpp::Rcout << "Rcpp vector is " << std::endl << x << std::endl;
}


// [[Rcpp::export]]
void callPrint(RObject x) { 
    Rcpp::print(x);             // will work on any SEXP object
}
```
```{r}
v <- seq(0.0, 10.0, by=2.5)
useOperatorOnVector(v)
callPrint(v)
```


## Be careful with pointers!
```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
NumericVector logRcpp1(NumericVector invec) {
    NumericVector outvec(invec);
    for(int i=0; i<invec.size(); i++) {
        outvec[i] = log(invec[i]);
    }
    return outvec;
}
```

```{r}
x <- seq(1.0, 3.0, by=1)
cbind(x, logRcpp1(x))
```
Note: outvec and invec point to the same underlying R object.

Instead, Use clone to not modify original vector.
```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
NumericVector logRcpp2(NumericVector invec) {
    NumericVector outvec=clone(invec);
    for(int i=0; i<invec.size(); i++) {
        outvec[i] = log(invec[i]);
    }
    return outvec;
}
```
```{r}
x <- seq(1.0, 3.0, by=1)
cbind(x, logRcpp2(x))
```

# RcppArmadillo

* Armadillo is a high level and easy to use C++ linear algebra library with
  syntax similar to Matlab.
* RcppArmadillo is an Rcpp interface allowing access to the Armadillo library.
* Supports a large body of linear algebra functions.

```{Rcpp, eval=FALSE}
#include <RcppArmadillo.h>

// [[Rcpp::depends(RcppArmadillo)]]

// [[Rcpp::export]]
void showMatrix(arma::mat X) {
    Rcout << "Armadillo matrix is" << std::endl << X << std::endl;
}

// [[Rcpp::export]]
List svd_arma(arma::mat x) {
    arma::mat xtx = x.t() * x ;
    arma::mat U ;
    arma::vec s ;
    arma::mat V ;
    svd(U, s, V, xtx) ;
    List ret ;
    ret["U"] = U ;
    ret["s"] = s ;
    ret["V"] = V ;
    return(ret) ;
}
```

```{r}
M <- matrix(1:9,3,3)
showMatrix(M)

## SVD implemented in Armadillo
svd_arma(M)
```
Note above was an example of returning multiple objects to R in a list.


# Do you need Rcpp?

Before using `Rcpp` in your package, you should first consider if it's
necessary.


## Typical R bottlenecks

* Loops that depend on previous iterations
  * MCMC methods, EM Algorithm, gradient descent...
* Function calls in R slow, but very little overhead in C++.
  * Recursive functions are very inefficient in R.
* Not having access to advanced data structures algorithms in R but available in C++.

## When is Rcpp not the solution?

* Sometimes the solution is to become a better R coder.
* Take advantage of vectorization when possible.
* Most base R functions already call C functions. Make sure there isn’t already
  an efficient implementation of what you are trying to do.
* If your bottleneck is with manipulating large rectangular data (*ie* over
  1 million rows or thousands of columns), consider <tt>data.table</tt>.

# Other resources

* vignette("Rcpp-quickref")
* `Seamless R and C++ integration with Rcpp' by Dirk Eddelbuettel. Excellent book for learning Rcpp. Available for free through JHU library.
* Hadley Wickham's Rcpp tutorial: http://adv-r.had.co.nz/Rcpp.html
* FAQ: https://cran.r-project.org/web/packages/Rcpp/vignettes/Rcpp-FAQ.pdf
* A huge number of examples at http://gallery.rcpp.org
* [Rcpp for Everyone](https://teuder.github.io/rcpp4everyone_en/)
* Stack exchange.
