---
title: "Skew normal mixture"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---



<style type="text/css">

body{ /* Normal  */
      font-size: 18px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 18px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Objectives

Normal mixture models are a popular method of distribution based clustering.
Despite this, in practice they often don't work well as slight departures
from normality lead to poor model fit. One way to remedy this is to use a more
robust distribution for clustering, such as adding additional parameters
for kurtosis (i.e. t-distribution) or skewness.

A finite dimensional mixture of skew-normal distributions assumes data
$y = (y_1, \ldots, y_n) \in R^n$ are a sample from a probability density function
of the form
$$
  f_{SN}(y; \xi, \omega^2, \alpha) = \frac{2}{\omega}
  \phi\left( \frac{y - \xi}{\omega}\right) \Phi(\alpha \omega^{-1}(y - \xi))
$$
where $\alpha$ is a skewness parameter.

Full conditionals are available for the proper parameter transformations and
Gibbs sampling is still feasible. See
'Bayesian inference for finite mixtures of univariate and multivariate skew-normal and skew-t distributions.'
by Früwirth-Schnatter, Pyne (2010) for derivations and greater detail.

See github.com/scristia/SkewNormalMix for an implementation of this model.

```{r load_packages, warning=FALSE, message=FALSE, results="hide", echo = TRUE}
# Load packages
library(tidyverse)
library(devtools)
#install_github("scristia/SkewNormalMix")
load_all("~/Software/SkewNormalMix")

library(sn)
library(mvtnorm)
library(msm)
library(MASS)
library(gtools)
library(truncnorm)
```

```{r simulate_data, echo=TRUE}
# simulated data
omega <- c(4, 1)
omega2 <- omega^2
alpha <- c(-3, 0)
mu <- c(0, 4)

xx <- c(rsn(5000, mu[1], omega[1], alpha[1]), rsn(8000, mu[2], omega[2], alpha[2]))
xx <- xx[sample.int(8000)]
par(bg="white")
plot(density(xx), type="l")

n <- length(xx)


##transformations
delta <- alpha/sqrt(1+alpha^2)
Ey <- mu+omega2*delta*sqrt(2/3.1415)
psi <- omega*delta
sigma2 <- omega2*(1-delta^2)
K = 2
# nsim=10000
nsim=2000
burnin <- 1:500
```

R implementation of function:

```{r, echo = TRUE, results="hide", warning=FALSE}
set.seed(4321)
res = skewnormal.gibbs(xx, K=K, nsim=nsim)
mus <- colMeans(res$MU[-burnin, ])
omegas <- colMeans(res$OMEGA[-burnin, ])
alphas <- colMeans(res$ALPHA[-burnin, ])
etas <- colMeans(res$ETA[-burnin, ])

```

Rcpp implementation of function:

```{r, echo = TRUE, results="hide", warning=FALSE}
set.seed(4321)
res2 <- skewNormalCpp(r=xx, K=K, nsim=nsim)
mus2 <- colMeans(res2$MU[-burnin, ])
omegas2 <- colMeans(res2$OMEGA[-burnin, ])
alphas2 <- colMeans(res2$ALPHA[-burnin, ])
etas2 <- colMeans(res2$ETA[-burnin, ])
```

Check posterior fit:

```{r, echo=FALSE}
lim <- range(xx, na.rm=TRUE)
par(las=1, mfrow=c(1,1), mar=c(4, 4, 4, 4), bg="white")
hist(xx, breaks = 500, col='lightgray', border='gray', freq=FALSE, main="",
     xlim=lim)
y2 <- seq(min(xx), max(xx), len=5000)
#post.dens <- pis[1]*dsn(y2, mus[1], omegas[1], alphas[1] ) + pis[2]*dsn(y2, mus[2], omegas[2], alphas[2])
#lines(y2, post.dens,lwd=2)
#mx <- max(post.dens)
for(k in 1:K) lines(y2, etas2[k]*dsn(y2, mus2[k], omegas2[k], alphas2[k] ), col="gray40", lty=2, lwd=2)
lines(y2, rowSums(sapply(1:K, function(x) etas2[x]*dsn(y2, mus2[x], omegas2[x], alphas2[x]))), col="skyblue3", lwd=2)
```

Check truth:

```{r, echo=FALSE}
knitr::kable(tibble("Parameter" = c("mu1", "mu2", "omega1", "omega2",
                           "alpha1", "alpha2", "eta1", "eta2"),
           "Truth" = c(mu, omega, alpha, c(8/13, 5/13)),
           "Estimate (R)" = round(c(mus, omegas, alphas, etas), 2),
           "Estimate (Rcpp)" = round(c(mus2, omegas2, alphas2, etas2),2)),
      digits=3)

```

## Benchmarking
```{r, results="hide"}
library(rbenchmark)
res <- benchmark(skewNormalCpp(r=xx, K=2, nsim=200),
          skewnormal.gibbs(xx, K=2, nsim=200, thin=1),
          replications=3)
```

```{r, echo=FALSE}
knitr::kable(res, digits=3)
```
